%!TEX root = ../main.tex
%******************************
%	 Chapter 2
%*****************************

\chapter{Addressing the mean-variability dependency in scRNA-Seq data}  

\graphicspath{{Chapter2/Figures/}}

\vfill

\begin{Abstract}
As shown above, cell-to-cell transcriptional variability in otherwise homogeneous cell populations plays crucial roles in immune activation and increases with age. Single-cell RNA sequencing can characterise this variability in a transcriptome-wide manner. However, technical variation and the confounding between variability and mean expression estimates hinders meaningful comparison of expression variability between cell populations. To address this problem, we introduce an analysis approach that extends the BASiCS statistical framework to derive a residual measure of variability that is not confounded by mean expression. This measure is used to test changes in variability in parallel to changes in mean expression on a gene-specific level transcriptome-wide.
With this method, we assessed changes in variability for genes responding to CD4\plus{} T cell activation and detect a synchronisation of biosynthetic machinery components. Furthermore, cytokines such as Il2 that support the activation of surrounding cells by paracrine signalling are heterogeneously up-regulated upon immune activation. When profiling more subtle transcriptional changes during CD4\plus{} T cell differentiation, we detect opposing patterns of changes in variability between \textit{Tbx21} and \textit{Cxcr5}, which are markers for Th1 and Tfh cells, indicating a delayed commitment processes throughout differentiation. 
\end{Abstract}

\vfill

\newpage

\vspace*{\fill}

\begin{Comment}
\textbf{Declaration} I worked on this project in close collaboration with Catalina Vallejos, Arianne Richard, Sylvia Richardson and John Marioni. In this project, I extended the existing Bayesian framework (BASiCS) to model scRNA-Seq data. Besides calculating the underlying math and programming the model, I helped re-writing the R package to be published in \href{https://bioconductor.org/packages/release/bioc/html/BASiCS.html}{Bioconductor}. Catalina Vallejos co-supervised me and introduced me to principles of Bayesian statistics. Arianne Richard helped with the interpretation of biological data and helped writing the manuscript. Sylvia Richardson provided statistical help on a part of the project that is not presented here. John Marioni co-supervised me and suggested analysis to be performed. Catalina Vallejos, John Marioni and I designed the study. Catalina Vallejos, John Marioni and I wrote the manuscript. I kindly thank Dominic Gr\"un for providing the smFISH data matched to mESC scRNA-Seq data. The study has been published as:\\

Nils Eling, Arianne C. Richard, Sylvia Richardson, John C. Marioni, Catalina A. Vallejos. Robust expression variability testing reveals heterogeneous T cell responses. \emph{Cell Systems}, In press, 2018 
\end{Comment}

\vspace*{\fill}

\newpage

% Include different main sections of the first chapter
\input{Chapter2/Chapt2_files/intro}
\newpage
\input{Chapter2/Chapt2_files/results}

