% **** Thesis Acknowledgements

\begin{acknowledgements}      

First and foremost, I want to thank my supervisor John Marioni for offering me to work in his research group. 
John took the risk in supervising me as a PhD student from a wet-lab background and introduced me to statistical analyses. 
I'm thankful for all his enthusiasm and support which guided me to successfully finishing my PhD in the most enjoyable way. 
I also want to thank Catalina Vallejos for supervising me and turning me into a "Bayesian statistician". 
Not only did she support my work but also became a close friend. \\

This work builds on excellent collaborations. 
I want to thank Duncan Odom not only for allowing me to collaborate with his lab but also for guiding me through important career-related processes (e.g.~paper writing). 
Christina Ernst supported large parts of this work with experiments, data interpretation and late nights of editing. 
I also want to thank Celia Martinez-Jimenez for large-scale experimental support and data interpretation which led to a successful publication. 
Maike de la Roche and Hung-Chang Chen also played a crucial role for publishing the final version of this paper. 
Furthermore, I thank Detlev Arendt and Kaia Achim for collaborating on an early project during my PhD. 
My thesis advisory committee members, Sarah Teichmann, Wolfgang Huber and Duncan Odom provided fruitful discussions that advanced my research over the last four years. 
I also thank the EMBL PhD programme for funding my research and the EMBL-European Bioinformatics Institute as well as the CRUK-Cambridge Institute for offering such a friendly work place.\\

All current and former lab members of the Marioni lab deserve a huge thanks for providing such a pleasant and stimulating work environment. 
I specifically thank: Aaron Lun for his statistical support, entertainment and being a BK lunch companion; Michael Morgan for discussing what "noise" really means; Arianne Richard for biological insights into the confusing world of T cells. 
I also thank Anna, Chris, Luis, Antonio, Konrad, Catalina, Liora, Jonny, Tom, Ximena, Christina, and Rebecca for all the fun times during the last four years. \\

A special thanks also goes to everyone who read and commented on this thesis: Christina Ernst, Jack Monahan, Hannah Meyer, Michael Morgan, Catalina Vallejos, and John Marioni. \\

I thank my examiners Chris Wallace and Ben Lehner for a truly enjoyable viva and the valuable discussions, which improved the presentation of this thesis.

\newpage
\thispagestyle{empty}
 
Importantly, I want to thank my friends from Cambridge, Heidelberg and Essen for their moral support, all the memories and keeping me sane during stressful times. 
Omar, thank you for distracting me with pizza and comedy nights, endless trips to the Blue Moon and for being a good friend. 
Hannah, Jack, Christina, Catalina, Dani, Lara, Julia, it was always fun going to formals, punting or trying to discover as many pubs in Cambridge as possible.\\

Sebastian and Ruben, thank you for listening to my complaints and helping me whenever needed, and together with Jonathan, Felix, Georg, Fabian for years of friendship, and for the most exhausting but fun "Alumni Wochenenden".  \\

I also want to thank Nico, Fabian, Josh, Martin, Marc, Alex, and Mike for always welcoming me back home. 
Even though this only happens few times per year, I truly value your friendship. \\

The Pembroke College Boat Club helped me to find a balance between sports and work and even during high-intensity phases I fully enjoyed the early mornings and the distraction that rowing offered. \\

Finally, I want to thank Bodi for her constant support and patience during the last four years. 
Our adventures, Skypes, holidays and weekends together gave me the energy to focus on my work.

\end{acknowledgements}
